{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-09-24 Docs/manuscript.Rnw
more edits, fixed plots

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{July 2023}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

% Just jotting down and putting some bits you have for abstract for now. 
\begin{abstract}
Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions. Yet we often lack this information entirely, or find conflicting evidence across studies. This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration. Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence. Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events. 
\end{abstract} %Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because...

\section{Introduction}
%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much? I suspect it may be contested. I also wonder if my advice about definitions was wrong ... but let's definitely save that text for now. 
%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.

@@

These contrasting studies highlight how inconsistent evidence for adaptation in tree spring phenology has been. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have failed to find similar trends along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent. %EMWSep4: Check deans96 is accurate ref I added. Other refs I added to 'more' bib in case you do not have them.

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) from 1960-1991 for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}. We verified our calculated MAT was was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.

@@

\section{Results}
Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species (five in North America and two in Europe) also had fall event information available. %EMWSep4 -- check my edits here; also if we don't have it somewhere in the results yet, please add gymnosperms/angiosperm species # in North America versus Europe here, something like: Most species in North American gardens were gymosperms (7/11 species) while most species were angiosperms in Europe (3/4 species).  
<<label=source_spring_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_spring_models_for_sweave.R")
@

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

%EMWSep4 -- small edits here, hopefully I did not screw up the formatting
In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more norther, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
    \caption{Posterior distributions for the effect of provenance latitude across different continents and species types. Zero---no effect---is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text). (A) Effect of latitude and continent on spring and fall event date (DOY). Fall events advanced strongly per degree northward, particularly notably in North America. (B) Effect of latitude and leaf type on spring and fall event date (DOY). As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms. Fall events advanced per degree northward for both species types.}
    \label{figure:continental_spptype_lat}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
    \caption{Posterior distributions for the effect of MAT across different continents and species types. Zero---no effect---is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text). (A) Effect of MAT and continent on spring and fall event date (DOY). Fall events advanced strongly with decreasing MAT, particularly notably in North America. (B) Effect of MAT and leaf type on spring and fall event date (DOY). As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms. Fall events delayed in warmer locations for both species types.}
    \label{figure:continental_spptype_effect}
\end{figure}

@@

Fall event clines were more similar across species types, but diverged by continent with the common clines of later events in warmer, more southerly latitudes found only in North American studies. However, as we were focused on spring events, we had limited data on fall events, which likely explains our results which contrast with meta-analyses of fall events in Europe (citeAlbero?). In our dataset, more data on fall events were reported in North America, while only two of the six European gardens reported leaf senescence-related metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}, and neither showed strong clines. In North America studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (EXPLAIN what their metric was?, cite Table S1). Budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
%EMWSep4 -- I am actually not sure if `our results which contrast with meta-analyses of fall events in Europe' is true -- is it? I could not tell from reviewing Alberto very quickly. If it is true from that paper or another, then I think this is the best explanation of our results. If it's not true, I can adjust the text.
% Also, can you say how the two gardens defined `leaf senescence' inside (EXPLAIN)?

%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph. I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper). 
Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult. Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations. In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation. Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies (CITESa). Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}). This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig. \ref{figure:map_gardens}). Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (CITES). Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly. 
%EMWSep4 -- Great point you made about the bias in North American study locations, I built on it. 
% CITESa -- could be Basler & Korner, Gauzere 2020 paper .. you likely have more ideas to add.

@@

%EMWSep4 -- good stuff here! I tried to refocus on spring events some as our data is focused there. And I limited the amount of hand-wavey stuff we discussed as reviewers only usually want so much extrapolation of your results (so I cut some of the ending stuff). 
Our results suggest that current advances in spring phenology will likely continue, but predicting any shifts due to provenance effects will require new data across more species. Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point. Beyond that point theory suggests plastic traits may limit adaptation to climate change (CITE chevin I think?). This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation (cite Gauzere 2020?).

@@

\section{Figures}

\bibliography{bibliography_local_adaptation.bib}

>>>>

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{2023}
\fancyhead[RO]{Manuscript}
\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it). However, I dont know how not to number the title page, and I also would like the number to appear bigger. Can lizzie fix this please?
\linenumbers
\usepackage{setspace}
% Using \doublespacing in the preamble changes the text to double-line spacing
\doublespacing
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\begin{abstract}
Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions. Yet we often lack this information entirely, or find conflicting evidence across studies. This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration. Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence. Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events. 
\end{abstract} %Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because...

\section{Introduction}

%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much? I suspect it may be contested. I also wonder if my advice about definitions was wrong ... but let's definitely save that text for now. 
%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.

@@

These contrasting studies highlight how inconsistent evidence for adaptation in tree spring phenology has been. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have failed to find similar trends along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent. %EMWSep4: Check deans96 is accurate ref I added. %alinaSep22: Hi lizzie I am not sure how you are trying to phrase "they may be more studies and debated in Europe" but I think it's slightly off. could you fix it?

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies. Note: map lines delineate study regions and do not necessarily depict accepted national boundaries} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) from 1960-1991 for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}. We verified our calculated MAT was was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.

@@

\section{Results}
Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species (five in North America and two in Europe) also had fall event information available. Most species in North American gardens were gymosperms (7/11 species) while most species in European gardens were angiosperms (3/4 species).

<<label=source_spring_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_spring_models_for_sweave.R")
@

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
    \caption{Posterior distributions (99\% percentile) for the effect of provenance latitude across different continents and species types. Zero---no effect---is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text). (a \& b) Effect of latitude and continent on spring and fall event date (DOY). Fall events advanced strongly per degree northward, particularly notably in North America. (c \& d) Effect of latitude and leaf type on spring and fall event date (DOY). As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms. Fall events advanced per degree northward for both species types.}
    \label{figure:continental_spptype_lat}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
    \caption{Posterior distributions (99\% percentile) for the effect of MAT across different continents and species types. Zero---no effect---is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text). (a \& b) Effect of MAT and continent on spring and fall event date (DOY). Fall events advanced strongly with decreasing MAT, particularly notably in North America. (c \& d) Effect of MAT and leaf type on spring and fall event date (DOY). As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms. Fall events delayed in warmer locations for both species types.}
    \label{figure:continental_spptype_effect}
\end{figure}

@@

In our dataset, more data on fall events were reported in North America. In Europe, only two gardens reported fall event metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}. Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}. North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (Supplement Table.\ref{supplement-table:all_studies}). Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree \citep{rosique22}. We note that budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.

%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph. I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper). 
Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult. Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations. In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation. Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}. Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}). This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig. \ref{figure:map_gardens}). Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g.,\emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}). Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly.

@@

%EMWSep4 -- good stuff here! I tried to refocus on spring events some as our data is focused there. And I limited the amount of hand-wavey stuff we discussed as reviewers only usually want so much extrapolation of your results (so I cut some of the ending stuff). 
Our results suggest that current advances in spring phenology will likely continue, but predicting any shifts due to provenance effects will require new data across more species. Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point. Beyond that point theory suggests plastic traits may limit adaptation to climate change \citep{chevin10}. This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation \citep{gauzere2020}.

@@

\section{Figures}
%alinaSep22: i dont know how to force all figures to be in this section instead of in between texts

\section{References}
\bibliography{bibliography_local_adaptation.bib}
